---
title: "Lab 4"
author: "Jazmin Hernandez"
format: html
editor: visual
embed-resources: true 
---

## 1. Read in the data

```{r}
met <- data.table::fread("met_all.gz")
```

## 2. Prepare the data

```{r}
met <- met[met$temp >= -17, ]
met[met$temp %in% c(9999, 999)] <- NA
met[met$wind.sp %in% c(9999, 999)] <- NA
met[met$dew.point %in% c(9999, 999)] <- NA
met[met$date %in% c(9999, 999)] <- NA
met[met$elev %in% c(9999, 999)] <- NA
met[met$wind.dir %in% c(9999, 999)] <- NA
```

```{r}
met$date <- as.Date(paste(met$year, met$month, met$day, sep = "-"), format = "%Y-%m-%d")
```

```{r}
library(data.table)
library(lubridate)
met[, week_of_year := week(date)]
met <- met[day(date) <= 7]
```

```{r}
mean_by_station <- met[, .(
  mean_temp = mean(temp, na.rm = TRUE),
  mean_rh = mean(rh, na.rm = TRUE),
  mean_wind_sp = mean(wind.sp, na.rm = TRUE),
  mean_vis_dist = mean(vis.dist, na.rm = TRUE),
  mean_dew_point = mean(dew.point, na.rm = TRUE),
  mean_lat = mean(lat, na.rm = TRUE),
  mean_lon = mean(lon, na.rm = TRUE),
  mean_elev = mean(elev, na.rm = TRUE)
), by = USAFID]
head(mean_by_station, 10)  # Displays the first 10 rows
```

```{r}
met[, region := fifelse(lat >= 39.71 & lon < -98, "NW",
                 fifelse(lat >= 39.71 & lon >= -98, "NE",
                 fifelse(lat < 39.71 & lon < -98, "SW", "SE")))]
```

```{r}
met[, elev_cat := fifelse(elev > 252, "high", "low")]
```

## 3. Use geom_violin to examine the wind speed and dew point by region

```{r}
library(ggplot2)
ggplot(na.omit(met), aes(x = factor(1), y = wind.sp)) +
  geom_violin(trim = FALSE) +
  facet_wrap (~region) +
  labs(x = "wind speed", y = "Density", title = "Geom_violin plot of Wind Speed by Region") +
  theme_minimal() +
  theme(axis.title.x = element_blank())
```

Looking at the violin plots, it seems that NE, NW, and SE regions have a pretty low density of wind speed values looking at the section that is the widest for the respective regions. However, the SW region does seem to have the most variation in wind speed values as we can see that there are more wider sections at varying density distributions.

```{r}
ggplot(na.omit(met), aes(x = factor(1), y = dew.point)) +
  geom_violin(trim = FALSE) +
  facet_wrap(~ region) +
  labs(x = "Dew Point", y = "Density", title = "Geom_violin Plot of Dew Point by Region") +
  theme_minimal() +
  theme(axis.title.x = element_blank())
```

In this plot, it appears that NE and SE regions have a similar dew point concentration at around the same density distribution. Again, the SW region seems to be the one that varies the most with most of its dew point concentration located at higher density and the next majority of the concentration is in lower density. In contrast, the NE and SE regions have a higher dew point concentration at higher density and higher wind speeds at lower densities.

## 4. Use geom_jitter with stat_smooth to examine the association between dew point and wind speed by region

```{r}
ggplot(na.omit(met), aes(x = dew.point, y = wind.sp, color = region)) +
  geom_jitter(alpha = 0.5, width = 0.2, height = 0.2) +  
  stat_smooth(method = "lm", se = TRUE) +  
  labs(x = "Dew Point", y = "Wind Speed", title = "Association between Dew Point and Wind Speed by Region")
```

There seems to be a slight negative correlation between wind speed and dew point looking at NW and SW regions. However, this differs from the NE and SE regions with a slightly positive association in dew point and wind speed. It appears that for the SE region, a higher dew point is found at a lower wind speed. On the contrary, a lower dew point for the SW region appear to be related to a higher wind speed. The difference in regression lines or association between the NW and SW regions with the SE and NE regions can be due to climatic effects and even the effect of ocean currents.

## 5. Use geom_bar to create barplots of the weather stations by elevation category colored by region

```{r}
ggplot(na.omit(met), aes(x = elev_cat, fill = region)) +
  geom_bar(position = "dodge") +  
  scale_fill_brewer(palette = "Accent") +  
  labs(x = "Elevation Category", y = "Weather Stations", title = "Weather Stations by Elevation Category and Region")
```

From the graph, it is clear that the SE region has the most weather stations in lower elevation categories while SW has the most weather stations in the higher elevation categories. The significant number of weather stations in the SE region located at a lower elevation can indicate that major climatic occurrences are of importance at this elevation in the SE region.

## 6. Use stat_summary to examine mean dew point and wind speed by region with standard deviation error bars

```{r}
ggplot(na.omit(met)) +
stat_summary(aes(x = region, y = dew.point), 
fun.data = "mean_sdl", geom = "bar", fill = "yellow") +
stat_summary(aes(x = region, y = dew.point), fun.data = "mean_sdl", geom = "errorbar") + 
stat_summary(aes(x = region, y = wind.sp), fun.data = "mean_sdl", geom = "bar", fill = "pink") +
stat_summary(aes(x = region, y = wind.sp), fun.data = "mean_sdl", geom = "errorbar") +
labs(x = "Region", y = "Value", title = "Mean Dew Point and Wind Speed by Region with Standard Deviation Bars")
```

The dew point and wind speed seem to be significantly different between NE and SE regions. However, there does appear to be an overlap in error bars between dew point and wind speed in the NW and SW regions. Dew point is higher in NE and SE regions. There appears to be the most variation in dew point and wind speed in the SW region shown by the larger error bars.

## 7. Make a map showing the spatial trend in relative humidity in the US

```{r}
library(leaflet)
library(RColorBrewer)
met_clean <- na.omit(met)
palette <- colorNumeric(palette = brewer.pal(11, "RdYlGn"), domain = met_clean$rh)
top_rh <- met_clean[order(-rh)][1:10]
leaflet(data = met_clean) |>
addTiles() |>
addCircles(
lng = ~lon, 
lat = ~lat, 
weight = 1, 
radius = 500,
color = ~palette(rh),  
stroke = TRUE, 
fillOpacity = 0.5) |>
addMarkers(
data = top_rh, 
lng = ~lon, 
lat = ~lat, 
popup = ~paste("USAFID:", USAFID)  ) |>
addLegend("bottomright", 
pal = palette, 
values = ~rh,
title = "Relative Humidity (%)",
opacity = 1)
```

Relative humidity seems to be the lowest in California and in Texas and in areas west in general. However, moving towards the east region, we can see that relative humidity increases, especially looking at the NE and SE regions.

## 8. Use a ggplot extension

```{r}
library(ggforce)
ggplot(met_clean, aes(x = temp, y = rh, color = elev_cat)) +
  geom_point(alpha = 0.6) +  
  geom_circle(aes(x0 = temp, y0 = rh, r = 2), alpha = 0.2) +  
  facet_wrap(~ elev_cat) +  
  labs(title = 'Relationship between Temperature and Relative Humidity',
       x = 'Temperature',
       y = 'Relative Humidity') +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
library(ggplot2)
library(patchwork)
p1 <- ggplot(met_clean) + geom_point(aes(rh, wind.sp))
p2 <- ggplot(met_clean) + geom_boxplot(aes(elev_cat, wind.sp, group = elev_cat))
p3 <- ggplot(met_clean) + geom_smooth(aes(dew.point, wind.sp))
p4 <- ggplot(met_clean) + geom_bar(aes(region))
(p1 | p2 | p3) /
      p4

```
